### 2020-07-28 CILAMCE_2020/paper.tex
Machine learning section and more figures

The methodology was carried out using RStudio software \cite{RStudio}, an integrated virtual environment for code development in R language \cite{RCore}. Throughout the process, all code executed was documented in the same order as its execution and pushed to the github repository (TODO: CREATE GITHUB REPO BIBLIO AND ADD HERE). The repository contains an extended version of this paper, including all the code, required packages and versions. In order to guarantee reproducibility, whenever there was code that uses probabilistic operations, a seed was defined before its execution, ensuring results consistency when running on another machine.

The main package used to build the machine learning models was the Caret Package \cite{caret}. It provides all functionalities and utilities to build prediction models for any data set, has a straight and clear documentation that guides the process and provide around 200 different algorithms to build models. We used only one algorithm in this work, chosen by the highest probability to achieve the best possible result, according to \citet{Fernandez2014}, who compared 179 algorithms across 121 different databases, and find out that the most likely to achieve the best possible results is the Parallel Random Forest (called prRF in the caret).

% --------------------------------------------------------------------------
\subsection{Dataset} \label{subsec:dataset}
% --------------------------------------------------------------------------
The data was downloaded from the University of California Irvine website \cite{downloadData}. The same dataset used by \citet{Yeh1998} and the related works. In total there are 1030 rows with 9 columns. Each row represents a sample with the variables: compressive strength, age, and 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers). TODO: referencia link do github com o arquivo xls utilizado (os dados originais)

@@

\item Ages of 90, 91, and 100 days were joined, transformed into 100 days. Three criteria were used to evaluate this step: First the plot of samples against a boxplot, that showed for example that most samples of 91 days have compressive strength between 60 and 80 MPa, while at the of age 90 days they are mostly between 30 and 50 MPa, but the difference of 1 day can't have such influence in the compressive strength. Then, because of the boxplot, a principal component analysis (PCA) of the ingredients were made, showing that the this fact is because the combination of ingredients among samples o 90 and 91 days are very distinct, same happens for 100 days. So it is very fair that we can join these ages without any prejudice to the predictions, in fact it can enhance the predictions by reducing noise. Finally, it was chosen to be 100 days because a plot of the compressive strength through time for IDs that have at least 5 different ages show that the compressive tends to increase through time, never decreases, as expected.

@@

After these manipulations, the final number of samples has been reduced from 1030 rows to 916 rows, with 416 different ingredient configurations (IDs). A xls file named "TODO" of the data at this point is available at the github repo (TODO: referencia bibliografica do link do arquivo no github, que tem o nome de "data\_2.xls", preciso fazer uma pasta com os arquivos e nomear de forma melhor no github - ainda vou fazer uma limpa no github e excluir muita coisa lá).

@@

Several plots and tables were built to perform the visualization of the samples prepared in the previous step. Including analysis of the descriptive statistics of the continuous and categorical variables, the distribution of the variables in relation to the compressive strength, the correlation between the variables grouped by age, the relationship between the approximate mix and the compressive strength, the relationships between the main concrete ingredients and the compressive strength and principal component analysis (PCA) of the ingredients. All of these plots/tables are part of the extended version available on github. (TODO: reference). However, to keep the article short, here only the plots of the statistical analysis of the categorical variables are presented, which are used for an insight in the conclusion of this work.

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.8]{Figures/stat-summ-categorical.pdf}
\vspace{12pt}
\caption{Descriptive statistics of the categorical variables}
\label{fig:categorical}
\end{center}
\end{figure}
\vspace{8pt}

@@

...

% --------------------------------------------------------------------------
\section{Results}\label{sec:Results}
% --------------------------------------------------------------------------
The test RMSE for each model in ascending order of age was 3.31, 4.36, 4.62, 4.72, 5.94 and 5.85 respectively. Table \ref{tab:results} presents the details and results of each model, including the naive one. Fig. \ref{fig:results} compares the actual and predicted values for the final models.

\vspace{8pt}
\begin{table}[!ht]
\centering
\caption{Final models}
\label{tab:results}
\vspace{6pt}
\begin{tabular}{ccccccc}
\hline
Model & mtry & CV & Repetitions & Naive RMSE (test) & Final RMSE (train) & Final RMSE (test)\\ \hline
3 days & 6 & 30 & 10 & 9.303229 & 3.905196 & 3.310370\\
7 days & 2 & 10 & 10 & 13.443646 & 4.475981 & 4.361987\\
14 days & 13 & 30 & 10 & 7.593319 & 5.136687 & 4.620515\\
28 days & 11 & 30 & 10 & 14.283824 & 5.847334 & 4.716698\\
56 days & 8 & 30 & 10 & 12.702112 & 6.702565 & 5.939163\\
100 days & 8 & 10 & 10 & 12.614652 & 6.381940 & 5.851088\\ \hline
\end{tabular}
\end{table}
\vspace{8pt}

>>>>

The methodology was carried out using RStudio software \cite{RStudio}, an integrated virtual environment for code development in R language \cite{RCore}. Throughout the process, all code executed was documented in the same order as its execution and pushed to the github repository (TODO: reference). The repository contains an extended version of this paper, including all the code, required packages and versions. In order to guarantee reproducibility, whenever there was code that uses probabilistic operations, a seed was defined before its execution, ensuring results consistency when running on another machine.

% --------------------------------------------------------------------------
\subsection{Dataset} \label{subsec:dataset}
% --------------------------------------------------------------------------
The data was downloaded from the University of California Irvine website \cite{downloadData}. The same dataset used by \citet{Yeh1998} and the related works. In total there are 1030 rows with 9 columns. Each row represents a sample with the variables: compressive strength, age, and 7 ingredients (water, cement, fine aggregate, coarse aggregate, fly ash, blast furnace slag, and superplasticizers). TODO: reference

@@

\item Ages of 90, 91, and 100 days were joined, transformed into 100 days. Three criteria were used to evaluate this step: First the plot of samples against a boxplot grouped by age, shown in Fig. \ref{fig:boxplot}, that showed for example that most samples of 91 days have compressive strength between 60 and 80 MPa, while at the of age 90 days they are mostly between 30 and 50 MPa, but the difference of 1 day can't have such influence in the compressive strength. Then, because of the boxplot, a principal component analysis (PCA) of the ingredients were made in Fig. \ref{fig:pca}, showing that the this fact is because the combination of ingredients among samples o 90 and 91 days are very distinct, same happens for 100 days. So it is very fair that we can join these ages without any prejudice to the predictions, in fact it can enhance the predictions by reducing noise. Finally, it was chosen to be 100 days because a plot of the compressive strength through time for IDs that have at least 5 different ages show that the compressive tends to increase through time, never decreases, as expected.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.7]{Figures/boxplot.pdf}
\vspace{12pt}
\caption{Boxplot - compressive strength grouped by age}
\label{fig:boxplot}
\end{center}
\end{figure}
\vspace{8pt}

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.7]{Figures/pca-90-91-100.pdf}
\vspace{12pt}
\caption{Principal component analysis - 90, 91 and 100 days}
\label{fig:pca}
\end{center}
\end{figure}
\vspace{8pt}

After these manipulations, the final number of samples has been reduced from 1030 rows to 916 rows, with 416 different ingredient configurations (IDs). A xls file named "TODO" of the data at this point is available at the github repo (TODO: referencia).

@@

Several plots and tables were built to perform the exploration and visualization of the samples prepared in the previous step. Including analysis of the descriptive statistics of the continuous and categorical variables, the distribution of the variables in relation to the compressive strength, the correlation between the variables grouped by age, the relationship between the approximate mix and the compressive strength, the relationships between the main concrete ingredients and the compressive strength and principal component analysis (PCA) of the ingredients. All of these plots/tables are part of the extended version available on github. (TODO: reference). However, to keep the article short, here only the plots of the statistical analysis of the categorical variables are presented, in the Fig. \ref{fig:categorical}, which are used for an insight in the conclusion of this work.

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.7]{Figures/stat-summ-categorical.pdf}
\vspace{12pt}
\caption{Descriptive statistics of the categorical variables}
\label{fig:categorical}
\end{center}
\end{figure}
\vspace{8pt}

% --------------------------------------------------------------------------
\subsection{Pre-processing and data split} \label{subsec:preprocessing}
% --------------------------------------------------------------------------

The main package used to build the machine learning models was the Caret Package \cite{caret}. It provides all functionalities and utilities to build prediction models for any data set, has a straight and clear documentation that guides the process and provide around 200 different algorithms to build models. In this work, it was done key steps described by \citet{irizarry2019} and \citet{Kuhn2008}. Starting by some pre-processing steps described below:

\begin{enumerate}

\item Removal of the categorical variables;

\item Separation of the dataset by age, resulting in 6 smaller datasets;

\item Removal of variables with near zero variance (only the fly ash of the 7-day set was removed);

\item Verification of variables with a correlation above 0.999, which did not occur;

\item Each data set was split into test and training sets, 20\% and 80\% respectively, shown in Table \ref{tab:split-config} and the distribution in the Fig. \ref{fig:split-dist}.

\end{enumerate}

\vspace{8pt}
\begin{table}[!ht]
\centering
\caption{Dataset split configurations}
\label{tab:split-config}
\vspace{6pt}
\begin{tabular}{cccc}
\hline
Model & Total samples & Train (80\%) & Test (20\%)\\ \hline
3 days & 121 & 97 & 24\\
7 days & 114 & 94 & 20\\
14 days & 62 & 50 & 12\\
28 days & 416 & 335 & 81\\
56 days & 83 & 67 & 16\\
100 days & 120 & 96 & 24\\ \hline
\end{tabular}
\end{table}
\vspace{8pt}

\begin{figure}[!ht]
\begin{center}
\includegraphics[scale=0.7]{Figures/dist-split.pdf}
\vspace{12pt}
\caption{Distribution of the test and train data in relation to compressive strength}
\label{fig:split-dist}
\end{center}
\end{figure}
\vspace{8pt}

% --------------------------------------------------------------------------
\subsection{Naive models} \label{subsec:naivemodels}
% --------------------------------------------------------------------------

Before building the real models, for comparison purposes, naive models were created. They simply predict that the compressive strength of the test set is the average compressive strength of the training set. In other words, naive models are simply the best guess possible to evaluate how close/far the real model is from a guess.

@@

We used only one algorithm in this work, chosen by the highest probability to achieve the best possible result, according to \citet{Fernandez2014}, who compared 179 algorithms across 121 different databases, and find out that the most likely to achieve the best possible results is the Parallel Random Forest (called prRF in the caret \cite{caret}). Six different models were built, one for each age-set, the following steps were made for each one:

\begin{enumerate}

\item Define the resampling scheme, with method of repeated cross validation;

\item Define a tuning grid for the "mtry" tuning parameter, which is a sequence from 1 to the number of columns of each dataset. All but the 7-day are equal since only the 7-day set have a column removed in the pre-processing;

\item Set seed equal to "1", chosen arbitrarily to guarantee reproducibility. This seed can be manipulated to obtain more satisfactory results, but it was chosen not to.

\item Do pre-processing transformation of the data with "center" and "scale" methods;

\item Run the caret "train" function with the above configurations and model "parRF";

\end{enumerate}

% --------------------------------------------------------------------------
\section{Results}\label{sec:Results}
% --------------------------------------------------------------------------
The performance evaluation of the models was performed by the Root Mean Square Error (RMSE). The RMSE is the measure used in all the works mentioned in the introduction allowing the comparison of the models in the discussion.

The test RMSE for each model in ascending order of age was 3.31, 4.36, 4.62, 4.72, 5.94 and 5.85 respectively. Table \ref{tab:results} presents the details and results of each model, including the naive one. Fig. \ref{fig:results} compares the actual and predicted values for the final models.

\vspace{8pt}
\begin{table}[!ht]
\centering
\caption{Final models results}
\label{tab:results}
\vspace{6pt}
\begin{tabular}{ccccccc}
\hline
Model & mtry & CV & Repetitions & Naive RMSE (test) & Final RMSE (train) & Final RMSE (test)\\ \hline
3 days & 6 & 30 & 10 & 9.303229 & 3.905196 & 3.310370\\
7 days & 2 & 10 & 10 & 13.443646 & 4.475981 & 4.361987\\
14 days & 13 & 30 & 10 & 7.593319 & 5.136687 & 4.620515\\
28 days & 11 & 30 & 10 & 14.283824 & 5.847334 & 4.716698\\
56 days & 8 & 30 & 10 & 12.702112 & 6.702565 & 5.939163\\
100 days & 8 & 10 & 10 & 12.614652 & 6.381940 & 5.851088\\ \hline
\end{tabular}
\end{table}
\vspace{8pt}
